---
title: "04-Amplification_bias_correction"
author: "Erin D'Agnese"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script is used after tourmaline-decontamination and ASV-taxon to correct reads based on amplification bias using the model and mock community sequencing. 

it requires:
1. QAQC of the mock community metadata - need to compile the metadata for all the Mocks then format the metadata like this but we only need to carry over the mass since for our mocks we made the mocks based on total DNA in anyway rather than needing to calc from ng/ul:

Species Barcode	Community	DNA_Ratio	DNA_conc_ng_uL	Total_DNA	
Citharichthys	sordidus	12S	Skew_Oceanic_1	6	18	108
Citharichthys	stigmaeus	12S	Skew_Oceanic_1	1	18	18
Citharichthys	xanthostigma	12S	Skew_Oceanic_1	1	18	18
Engraulis	mordax	12S	Skew_Oceanic_1	1	18	18
Leuroglossus	stilbius	12S	Skew_Oceanic_1	1	18	18
Merluccius	productus	12S	Skew_Oceanic_1	1	18	18
Nannobrachium	ritteri	12S	Skew_Oceanic_1	1	18	18
Sardinops	sagax	12S	Skew_Oceanic_1	6	18	108

2. QAQC mock sequence data so that we can join them to create a mock_community_data.rds that looks like this:
# A tibble: 6 × 6
  ID_mifish community    tech_rep Cycles nReads start_conc_ng
  <chr>     <chr>        <chr>    <chr>   <dbl>         <int>
1 Agonidae  Coastal_Even 1        30          0             0
2 Agonidae  Coastal_Even 1        33          0             0
3 Agonidae  Coastal_Even 1        36          0             0
4 Agonidae  Coastal_Even 1        39          0             0
5 Agonidae  Coastal_Even 2        30          0             0
6 Agonidae  Coastal_Even 2        33          0             0

3. the mock community model calibration using mock_community_data.rds and 

(For 18S and MiFish it was 39 cycles including the 14 TD cycles)

Notes from Zack:
We need to put in sanity checks to look at the stacked barplots of proportions 

For big skate and ratfish - if there are things in one rep of the mock but not in the others then they need to be removed 

In the Mock: Take the species that showed up in all the technical reps only and throw them out cuz it breaks the model 
Make a mock for the shrimp - take the alpha at the end and hard code

Filter observed for those species - drop any samples that have low reads (ZG did <5000 for 5 species) 

Run the TM and Ctrl as observed - may need to separate them, but shouldn't need to

Plot what they look like before and after for both mock and obsevered to make sure nothing weird is happening. 

Try it step-wise adding rather than removing - pare the mock down to the most important, get the model to run and then add more species in until it stops work


First let's load the packages for formatting the mock community data 
```{r}
library(here)
library(tidyverse)
```

load in the metadata for the mock community data
```{r}
mock_data <- read.csv(here("input-files","Mock_communities_18S_metadata.csv"), header=T, check.names = FALSE)
```
There is more columns than we need so let's clean that dataframe up - the start_conc_ng in Zack's input is actual total DNA so will just use the mass_ng since they are the same type of "conc"
```{r}
mock_data %>% 
  rename(taxon_assign = "18S_taxon_assign") %>% 
  rename(species = "Species") %>% 
  select(species,taxon_assign, Community, Even_prop, Even_mass_ng, SK1_prop, SK1_mass_ng, SK2_prop, SK2_mass_ng) -> mock_data

```

we need to make it longform with a specific name for each community that matches the samplenames in the sequencing metadata
```{r}
library(tidyverse)

long_mock_data <- mock_data %>%
  # Select relevant columns (excluding species name)
  select(taxon_assign, Community, ends_with("_mass_ng")) %>%
  # Pivot to long format
  pivot_longer(
    cols = ends_with("_mass_ng"),
    names_to = "sample_type",
    values_to = "start_conc_ng"
  ) %>%
  # Clean up sample type names and create community IDs
  mutate(
    sample_type = str_remove(sample_type, "_mass_ng"),
    community = paste(Community, sample_type, sep = "_")
  ) %>%
  rename(
    ID_18S = taxon_assign
  ) %>%
  # Remove original Community column (now in community)
  select(-Community) %>%
  # Group by ID_18S and community to sum concentrations
  group_by(ID_18S, community) %>%
  summarise(
    start_conc_ng = sum(start_conc_ng, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Remove any zero-concentration entries
  filter(start_conc_ng > 0) %>%
  # Sort for clean output
  arrange(community, ID_18S)

# First fix the naming in long_mock_data
long_mock_data <- long_mock_data %>%
  mutate(
    community = case_when(
      str_detect(community, "_SK1") ~ str_replace(community, "_SK1", "_Skew_1"),
      str_detect(community, "_SK2") ~ str_replace(community, "_SK2", "_Skew_2"),
      TRUE ~ community  # Keep other names unchanged
    )
  )

# View result
head(long_mock_data)
```

okay wee need the Mock community sequencing data now and we need to modify the names of the collumns and clean up the rows and taxonomy
```{r}
asv_data <- read.csv(here("intermediate-files","CorrectionBiasRun","CorBias_18S_taxon_decontam_asv_table.csv"), header = T, check.names = F)

asv_meta <- read.table(here("data","CorrectionBias","tourmaline-CorBias_18S_16072025","00-data","metadata.tsv"), sep = "\t", header = T)

asv_wild_data <- read.csv(here("intermediate-files","Wild_CB_samples","CorBias_18S_taxon_decontam_asv_table.csv"), header = T, check.names = F)

```



This is the older version for a different project, we are modifying directly in this script for our use
```{r}
#source(here("scripts","calibrate_metabarcoding.R"))
```



let's select out the mock community samples 
```{r}
asv_meta %>% 
  filter(region == "MockCom") -> mock_asv_meta

# First, get the sample IDs you want to keep from the metadata
samples_to_keep <- mock_asv_meta$sampleid

# Convert the ASV table to long format and filter for desired samples
filtered_asv_data <- asv_data %>%
  # Keep the taxonomy column
  select(final_taxonomy, all_of(samples_to_keep)) %>%
  # Convert to long format
  pivot_longer(
    cols = -final_taxonomy,
    names_to = "sampleid",
    values_to = "count"
  ) %>%
  # Filter to only include samples in your metadata
  filter(sampleid %in% samples_to_keep)

# The result is a long-format dataframe with taxonomy, sampleid, and count columns
head(filtered_asv_data)

mock_asv_data <- filtered_asv_data %>%
  left_join(mock_asv_meta, by = "sampleid")

```

now we need to clean up the data based on the known input species and remove any contaminants that are secondary prey from the species that we homogenized
```{r}
mock_asv_data <- mock_asv_data %>%
  # First separate samplename into all components
  separate(
    samplename,
    into = c("sampletype", "marker", "commtype", "commmix", "tech"),
    sep = "_",
    remove = FALSE
  ) %>%
  # Clean up technical replicates
  mutate(
    tech = as.integer(str_remove(tech, "^R")),  # Remove "R" and convert to integer
    # Reconstruct community name without tech rep
    community = paste(sampletype, marker, commtype, commmix, sep = "_")
  )

mock_asv_data <- mock_asv_data %>%
  # Standardize Skew naming in all relevant columns
  mutate(
    across(c(community, commmix, samplename),
    ~case_when(
      str_detect(.x, "Skew1") ~ str_replace(.x, "Skew1", "Skew_1"),
      str_detect(.x, "Skew2") ~ str_replace(.x, "Skew2", "Skew_2"),
      TRUE ~ .x
    )
  )) %>%
  # Final cleanup
  select(-sampletype, -marker) %>%  # Remove temporary columns if not needed
  # Filter for present taxa
  group_by(final_taxonomy) %>% 
  filter(any(count > 0)) %>%
  ungroup()

# First get all unique ID_18S values we need to match
ids_to_match <- unique(long_mock_data$ID_18S)

# Create a list of taxonomy strings that contain any ID_18S
valid_taxa <- mock_asv_data %>%
  filter(str_detect(final_taxonomy, paste(ids_to_match, collapse = "|"))) %>%
  pull(final_taxonomy) %>%
  unique()

# Filter and merge
merged_data_rep_fish <- mock_asv_data %>%
  filter(final_taxonomy %in% valid_taxa) %>%
  inner_join(long_mock_data, by = "community") %>%
  filter(str_detect(final_taxonomy, ID_18S)) %>%
  filter(commtype == "RepFish") %>% 
  mutate(
    community = paste(commtype, commmix, sep = "_")  # Combine commtype and commmix
  )

merged_data_comb_fish <- mock_asv_data %>%
  filter(final_taxonomy %in% valid_taxa) %>%
  inner_join(long_mock_data, by = "community") %>%
  filter(str_detect(final_taxonomy, ID_18S)) %>%
  filter(commtype == "CombFish") %>% 
  mutate(
    community = paste(commtype, commmix, sep = "_")  # Combine commtype and commmix
  )

```
this removed the ratfish because nothing assigned to chimaeriformes in the 18S mocks - there is not a good 18S reference seq for ratfish and apparently it is different enough from the other chondricthyes that it won't even assign to that clade.

Let's do a sanity check- barplot of existing props of all species in each mock
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

plot_taxonomy_proportions <- function(data, 
                                     taxonomy_col = "ID_18S",
                                     count_col = "count",
                                     group_cols = c("community", "tech"),
                                     top_n_taxa = 12,
                                     plot_title = "Taxonomic Proportions by Group") {
  
  # Calculate proportions
  prop_data <- data %>%
    group_by(across(all_of(c(group_cols, taxonomy_col)))) %>%
    summarise(total_count = sum(.data[[count_col]]), .groups = "drop_last") %>%
    mutate(proportion = total_count / sum(total_count)) %>%
    ungroup()
  
  # Get top taxa for each group
  top_taxa <- prop_data %>%
    group_by(.data[[taxonomy_col]]) %>%
    summarise(avg_prop = mean(proportion)) %>%
    arrange(desc(avg_prop)) %>%
    slice_head(n = top_n_taxa) %>%
    pull(.data[[taxonomy_col]])
  
  # Filter and order data
  plot_data <- prop_data %>%
    mutate(!!sym(taxonomy_col) := ifelse(.data[[taxonomy_col]] %in% top_taxa, 
                                        .data[[taxonomy_col]], "Other"),
           !!sym(taxonomy_col) := factor(.data[[taxonomy_col]], 
                                        levels = c(top_taxa, "Other"))) %>%
    group_by(across(all_of(c(group_cols, taxonomy_col)))) %>%
    summarise(proportion = sum(proportion), .groups = "drop")
  
  # Create combined grouping variable for plotting
  plot_data <- plot_data %>%
    unite("group", all_of(group_cols), sep = " | ")
  
  # Generate plot
  p <- ggplot(plot_data, aes(x = group, y = proportion, fill = .data[[taxonomy_col]])) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = plot_title,
         x = paste(group_cols, collapse = " | "),
         y = "Proportion of Counts",
         fill = taxonomy_col) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 7)) +
    scale_y_continuous(labels = scales::percent)
  
  if (length(top_taxa) > 5) {
    p <- p + guides(fill = guide_legend(ncol = 2))
  }
  
  return(p)
}

# Example usage:
p1 <- plot_taxonomy_proportions(merged_data_comb_fish, 
                               plot_title = "Initial Mock Check")

print(p1)
```
Okay that looks pretty normal for the combined fish mocks

Now the RepFish one
```{r}
p2 <- plot_taxonomy_proportions(merged_data_rep_fish, 
                               plot_title = "Initial Mock Check")

print(p2)
```

The rep fish also looks appropriate based on what we expect for the Rep fish mocks.

Reference these when comparing the versions that are transformed along the way to make sure the numbers don't get changed incorrectly.

This is all 12 of the taxons that are in the mocks.

Now we need to save the cleaned up mock data as an .RDS so we can bring it in for the mock calibration step
```{r}
#Using the RepFish mock first

merged_data_rep_fish %>% 
  select(ID_18S, community, tech, count,start_conc_ng) %>% 
  mutate(Cycles = 39) %>% 
  rename("nReads" = "count") -> mock_18S_RF_input

input_mock_RFcomm_RDS <- mock_18S_RF_input %>% 
  filter(., Cycles ==39) %>% 
  group_by(community, tech) %>% 
  mutate(., totReads=sum(nReads),
         propReads = nReads/totReads,
         tot_start = sum(start_conc_ng),
         start_prop=start_conc_ng/tot_start) %>% 
  dplyr::select(Community=community, Tech_Rep=tech,species=ID_18S, nReads,totReads,propReads,start_prop, N_pcr_mock=Cycles) %>% 
  mutate(Tech_Rep = as.numeric(Tech_Rep),
         N_pcr_mock = as.numeric(N_pcr_mock)) 

input_mock_RFcomm_RDS %>% 
  rename(tech = Tech_Rep) %>% 
  group_by(Community) %>%
  mutate(station_idx = cur_group_id()) %>%
  ungroup() %>% 
  mutate(station = station_idx) -> input_mock_RFcomm_RDS

#save for future use
saveRDS(input_mock_RFcomm_RDS, here("intermediate-files","Prey18S_RepFish_mock_community_data.RDS") )

#and now the CombFish
merged_data_comb_fish %>% 
  select(ID_18S, community, tech, count,start_conc_ng) %>% 
  mutate(Cycles = 39) %>% 
  rename("nReads" = "count") -> mock_18S_CF_input

input_mock_CFcomm_RDS <- mock_18S_CF_input %>% 
  filter(., Cycles ==39) %>% 
  group_by(community, tech) %>% 
  mutate(., totReads=sum(nReads),
         propReads = nReads/totReads,
         tot_start = sum(start_conc_ng),
         start_prop=start_conc_ng/tot_start) %>% 
  dplyr::select(Community=community, Tech_Rep=tech,species=ID_18S, nReads,totReads,propReads,start_prop, N_pcr_mock=Cycles) %>% 
  mutate(Tech_Rep = as.numeric(Tech_Rep),
         N_pcr_mock = as.numeric(N_pcr_mock)) 

input_mock_CFcomm_RDS %>% 
  rename(tech = Tech_Rep) %>% 
  group_by(Community) %>%
  mutate(station_idx = cur_group_id()) %>%
  ungroup() %>% 
  mutate(station = station_idx) -> input_mock_CFcomm_RDS

#save for future use
saveRDS(input_mock_CFcomm_RDS, here("intermediate-files","Prey18S_CombFish_mock_community_data.RDS") )

```

Now the mock data should be ready to input in the model calibration, but let's make sure the prop calc outputs worked as expected compared to the previous figure 
```{r}
plot_proportion_comparison <- function(data, 
                                      taxonomy_col = "species",
                                      count_col = "nReads",
                                      existing_prop_col = "propReads",
                                      group_cols = c("Community", "tech"),
                                      top_n_taxa = 12,
                                      plot_title = "Proportion Comparison") {
  
  # Calculate proportions from counts
  calc_data <- data %>%
    group_by(across(all_of(c(group_cols, taxonomy_col)))) %>%
    summarise(total_count = sum(.data[[count_col]]), .groups = "drop_last") %>%
    mutate(calculated_prop = total_count / sum(total_count)) %>%
    ungroup()
  
  # Get existing proportions if column exists
  if (existing_prop_col %in% names(data)) {
    existing_data <- data %>%
      group_by(across(all_of(c(group_cols, taxonomy_col)))) %>%
      summarise(existing_prop = sum(.data[[existing_prop_col]]), .groups = "drop")
    
    # Combine both datasets
    plot_data <- full_join(calc_data, existing_data, 
                          by = c(group_cols, taxonomy_col))
  } else {
    message("No existing proportion column found - plotting only calculated proportions")
    plot_data <- calc_data %>%
      mutate(existing_prop = NA_real_)
  }
  
  # Get top taxa based on calculated proportions
  top_taxa <- calc_data %>%
    group_by(.data[[taxonomy_col]]) %>%
    summarise(avg_prop = mean(calculated_prop)) %>%
    arrange(desc(avg_prop)) %>%
    slice_head(n = top_n_taxa) %>%
    pull(.data[[taxonomy_col]])
  
  # Prepare data for plotting
  plot_data <- plot_data %>%
    mutate(!!sym(taxonomy_col) := ifelse(.data[[taxonomy_col]] %in% top_taxa, 
                                        .data[[taxonomy_col]], "Other"),
           !!sym(taxonomy_col) := factor(.data[[taxonomy_col]], 
                                        levels = c(top_taxa, "Other"))) %>%
    group_by(across(all_of(c(group_cols, taxonomy_col)))) %>%
    summarise(calculated_prop = sum(calculated_prop, na.rm = TRUE),
              existing_prop = sum(existing_prop, na.rm = TRUE),
              .groups = "drop") %>%
    pivot_longer(cols = c(calculated_prop, existing_prop),
                 names_to = "proportion_type",
                 values_to = "proportion") %>%
    unite("group", all_of(group_cols), sep = " | ") %>%
    mutate(proportion_type = factor(proportion_type,
                                   levels = c("calculated_prop", "existing_prop"),
                                   labels = c("From Counts", "Existing Proportions")))
  
  # Generate plot
  p <- ggplot(plot_data, aes(x = group, y = proportion, fill = .data[[taxonomy_col]])) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~ proportion_type, ncol = 2) +
    labs(title = plot_title,
         x = paste(group_cols, collapse = " | "),
         y = "Proportion",
         fill = taxonomy_col) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 7),
          strip.text = element_text(face = "bold")) +
    scale_y_continuous(labels = scales::percent)
  
  if (length(top_taxa) > 5) {
    p <- p + guides(fill = guide_legend(ncol = 2))
  }
  
  return(p)
}

# Example usage:
p3 <- plot_proportion_comparison(input_mock_RFcomm_RDS,existing_prop_col = "propReads",
                                 plot_title = "Count vs Existing Proportions")
print(p3)
```
They look correct

```{r}
p4 <- plot_proportion_comparison(input_mock_CFcomm_RDS,existing_prop_col = "propReads",
                                 plot_title = "Count vs Existing Proportions")
print(p4)
```
The combined fish also look the same, so we can believe we are sane still.

Okay now we need to clean up the ASV tables of real-world sequence data. We will use the tissue mixture samples, the controlled feeding study and we will eventually pull some commencement bay data as wild samples to ensure it works across all the sample types. 
```{r}
asv_meta %>% 
  filter(str_detect(Plate_ID,"Prey18S")) %>% 
  filter(region %in% c("ControlFeeding","TissueMix")) -> asv_meta_18S_samples

# First, get the sample IDs you want to keep from the metadata
samples_to_keep <- asv_meta_18S_samples$sampleid
# Find which samples_to_keep actually exist in asv_data
existing_samples <- intersect(samples_to_keep, names(asv_data))

# Convert the ASV table to long format and filter for desired samples
filtered_asv_data <- asv_data %>%
  # Keep the taxonomy column
  select(final_taxonomy, all_of(existing_samples)) %>%
  # Convert to long format
  pivot_longer(
    cols = -final_taxonomy,
    names_to = "sampleid",
    values_to = "count"
  ) %>%
  # Filter to only include samples in your metadata
  filter(sampleid %in% existing_samples)

# The result is a long-format dataframe with taxonomy, sampleid, and count columns
head(filtered_asv_data)

samples18S_asv_data <- filtered_asv_data %>%
  left_join(asv_meta_18S_samples, by = "sampleid") %>% 
  select(samplename,region,final_taxonomy,count)

#now clean up data to only include taxa that are in the mock or related to things in the mock
# Function to parse sample names based on region
library(tidyverse)
library(stringr)

# Enhanced parser that handles complex station names
parse_sample_name <- function(samplename, region) {
  parts <- str_split(samplename, "_")[[1]]
  
  if (region == "ControlFeeding") {
    # Format: PvCF_D1_01_TR1 (exactly 4 parts)
    creek <- parts[1]
    station <- parts[2]
    biol <- as.integer(parts[3])
    tech <- as.integer(str_remove(parts[4], "TR"))
  } else if (region == "TissueMix") {
    creek <- "TM"
    biol <- as.integer(parts[2])
    
    # Check if last part is a tech rep (digits or R+digits)
    last_part <- tail(parts, 1)
    if (str_detect(last_part, "^(R?\\d+)$")) {
      tech <- as.integer(str_remove(last_part, "R"))
      # Station is everything between biol and tech
      station <- paste(parts[3:(length(parts)-1)], collapse = "_")
    } else {
      tech <- 1
      # Station is everything after biol
      station <- paste(parts[3:length(parts)], collapse = "_")
    }
  } else {
    return(list(creek = NA, station = NA, biol = NA, tech = NA))
  }
  
  return(list(creek = creek, station = station, biol = biol, tech = tech))
}


# Final transformation function
modified_sample_data <- function(samples18S_asv_data) {
  samples18S_asv_data %>%
    filter(region %in% c("ControlFeeding", "TissueMix")) %>%
    rowwise() %>%
    mutate(
      parsed = list(parse_sample_name(samplename, region)),
      creek = parsed$creek,
      station = parsed$station,
      biol = parsed$biol,
      tech = parsed$tech,
      time = 1,
      species = final_taxonomy,
      Nreads = count
    ) %>%
    ungroup() %>%
    # Keep samplename in the output for verification
    select(samplename, time, creek, station, biol, tech, species, Nreads) %>%
    filter(!is.na(creek))  # Remove any unparseable samples
}


# Run the transformation
sample_data <- modified_sample_data(samples18S_asv_data)

# Verify the transformation
head(sample_data)


```

We need to modify the taxonomy to be "species" to match the mock data
```{r}
extract_final_taxon <- function(full_taxonomy) {
  # Split by semicolon and remove empty elements
  taxa_levels <- str_split(full_taxonomy, ";")[[1]] %>% 
    str_trim() %>% 
    .[. != ""]
  
  # Work backwards through the taxonomy levels
  for (i in length(taxa_levels):1) {
    current_taxon <- taxa_levels[i]
    # Skip if the taxon is NA or empty
    if (!is.na(current_taxon) && current_taxon != "NA" && current_taxon != "") {
      return(current_taxon)
    }
  }
  
  # If all levels are NA/empty, return NA
  return(NA_character_)
}

# Process observation data to get proper species names
preprocess_observation_data <- function(observation_data) {
  observation_data %>%
    rename(full_tax = species) %>%
    mutate(
      species = map_chr(full_tax, extract_final_taxon),
      species = str_trim(species)
    ) %>%
    filter(!is.na(species), species != "") %>%
    select(-full_tax)
}

# Improved transformation pipeline
sample_data_modified <- sample_data %>%
  rename(full_tax = species) %>%
  mutate(
    species = map_chr(full_tax, extract_final_taxon),
    species = str_trim(species)
  ) %>%
  # Remove any rows where we couldn't extract a valid taxon
  filter(!is.na(species), species != "") %>%
  # Reorder columns
  select(samplename, time, creek, station, biol, tech, species, Nreads, full_tax)

# Verify the transformation
head(sample_data_modified)

# Check for any remaining NA/empty values in species
sum(is.na(sample_data_modified$species) | sample_data_modified$species == "")

# Check species names against mock data
species_in_samples <- unique(sample_data_modified$species)
species_in_mock <- unique(mock_18S_CF_input$ID_18S)
#species_in_mock <- unique(mock_18S_RF_input$ID_18S)

# See which species are in both datasets
intersect(species_in_samples, species_in_mock)

# See which species are only in samples (might need manual checking)
setdiff(species_in_samples, species_in_mock)

# After your existing species checks:
sample_data_modified <- sample_data_modified %>% 
  filter(species %in% species_in_mock)


sample_data_modified %>% 
  select(!full_tax) -> sample_data

#save for future use
saveRDS(sample_data, here("intermediate-files","Prey18S_CF_TM_input_combfishMC_data.RDS"))

```

okay let's split them into ControlFeeding study samples and TM samples in case we need to run them separately
```{r}
sample_data %>% 
  filter(creek == "PvCF") -> sample_data_PvCF

sample_data %>% 
  filter(creek == "TM") -> sample_data_TM
```


Create the additional mock for the Penaeus in the controlled feeding study
```{r}
#We need to see if we can include the Penaeus shrimp in the mocks some how but using the TM
# For the special TM samples to use as additional mock:
# Create additional mock community from specific TM samples
additional_mock <- sample_data_TM %>%
  filter(station %in% c("BlkTigShrVAQ", "VAQShrimp")) %>%
  mutate(
    # Create Community name by combining TM prefix with station
    Community = paste("TM", station, sep = "_"),
    # Convert biol to Tech_Rep (assuming biol should be the tech rep for mocks)
    tech = as.numeric(biol),
    # Set known proportions
    start_conc_ng = case_when(
      species == "Penaeus monodon" ~ 0.5,
      species == "Actinopteri" ~ 0.5,
      TRUE ~ 0
    )
  ) %>%
  dplyr::rename(nReads = "Nreads") %>% 
  filter(start_conc_ng > 0) %>%  # Remove species not in the mock design
  group_by(Community, tech) %>%
  mutate(N_pcr_mock = 39) %>% 
  mutate(
    # Calculate derived values to match mock structure
    totReads = sum(nReads),
    propReads = nReads / totReads,
    tot_start = sum(start_conc_ng),
    start_prop = start_conc_ng / tot_start
  ) %>%
  ungroup() %>%
  # Select and rename columns to match mock input structure
  dplyr::select(
    Community,
    tech,
    species,
    nReads,
    totReads,
    propReads,
    start_prop) %>%
  # Ensure no duplicates
  distinct()

#keep it separate to run and extract the alpha for Penaeus to add after the model is created to the rest of the models

input_mock_commTM_RDS <- additional_mock %>% 
  mutate(N_pcr_mock = 39) %>% 
  mutate(station = 1)

saveRDS(input_mock_commTM_RDS, here("intermediate-files","Prey18S_TMPenaeusMock_data.RDS"))

```


MODIFIED FROM ZG/EILY/OLE
```{r}
library(tidyverse)
library(here)
# Prepare for stan model 
input_metabarcoding_RDS <- sample_data %>% dplyr::select(!samplename)

input_mock_comm_RDS <- input_mock_CFcomm_RDS
#input_mock_comm_RDS <- input_mock_RFcomm_RDS

```

Attempt at formatting function mod 1
```{r}
format_metabarcoding_data <- function(input_metabarcoding_RDS, input_mock_comm_RDS) {
  require(tidyverse)
  
  Observation <- input_metabarcoding_RDS 
  
  # Create mock community to station index mapping
  mock_community_index <- input_mock_comm_RDS %>%
    distinct(Community) %>%
    mutate(station_idx = row_number())
  
  # Process Mock data
  Mock <- input_mock_comm_RDS %>% 
    left_join(mock_community_index, by = "Community", relationship = "many-to-many") %>%
    mutate(
      time = 1,    # for formatting compatibility
      creek = 1,
      biol = 1,
      station = station_idx,
      stationname = Community,
      Nreads = nReads  # Rename here to match Observation
    ) %>% 
    replace(is.na(.), 0) %>% 
    filter(start_prop > 0 & Nreads > 0 & !is.na(Nreads)) %>% 
    unite(c(time, creek, station, biol, tech), col = "Sample", sep = ".", remove = FALSE)
  
  # Keep species present in both datasets plus Penaeus monodon
  keepSpecies <- unique(c(intersect(Mock$species, Observation$species), "Penaeus monodon"))
  
  Observation <- Observation %>% filter(species %in% keepSpecies)
  Mock <- Mock %>% filter(species %in% keepSpecies) 

  # Create species index
  sp_list <- data.frame(
    species = unique(c(Mock$species, Observation$species)),
    species_idx = match(unique(c(Mock$species, Observation$species)), 
                      unique(c(Mock$species, Observation$species)))
  )
  
  # Process Mock data
  Mock <- Mock %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      speciesname = species,
      species = species_idx
    ) %>% 
    group_by(station, tech, time, creek, biol) %>% 
    mutate(b_proportion = start_prop/sum(start_prop)) %>% 
    ungroup()
  
  # Process Observation data
  Observation <- Observation %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      stationname = station,
      speciesname = species,
      species = species_idx
    ) %>% 
    mutate(station = case_when(
      stationname == "D1" ~ 1, stationname == "D2" ~ 2, stationname == "D3" ~ 3,
      stationname == "D4" ~ 4, stationname == "D5" ~ 5, stationname == "D6" ~ 6,
      stationname == "B_Skate" ~ 7, stationname == "MgArmSq" ~ 8, stationname == "BayShr" ~ 9,
      stationname == "MktSq" ~ 10, stationname == "GPO" ~ 11, stationname == "Ratfish" ~ 12,
      stationname == "RedOcto" ~ 13, stationname == "CoonStShrp" ~ 14, stationname == "SptPrawn" ~ 15,
      stationname == "PnkShrmp" ~ 16, stationname == "SpDog" ~ 17, stationname == "Lnose_skate" ~ 18,
      stationname == "RivLamp" ~ 19, stationname == "SqVAQ" ~ 20, stationname == "GhShrmp" ~ 21,
      stationname == "StubSq" ~ 22, stationname == "MusPolyWorm" ~ 23, 
      stationname == "BlkTigShrVAQ" ~ 24, stationname == "VAQShrimp" ~ 25,
      TRUE ~ NA_integer_
    )) %>% 
    group_by(station, time, creek, biol) %>% 
    mutate(
      tech = match(tech, unique(tech)),
      tot_reads = sum(Nreads)
    ) %>% 
    filter(tot_reads > 0) %>% 
    ungroup() %>% 
    dplyr::select(-tot_reads)
  
  # Create station metadata
  station_list <- data.frame(
    station = 1:25,
    stationname = c(paste0("D", 1:6),
                   "B_Skate", "MgArmSq", "BayShr", "MktSq", "GPO", 
                   "Ratfish", "RedOcto", "CoonStShrp", "SptPrawn",
                   "PnkShrmp", "SpDog", "Lnose_skate", "RivLamp",
                   "SqVAQ", "GhShrmp", "StubSq", "MusPolyWorm",
                   "BlkTigShrVAQ", "VAQShrimp"),
    station_type = c(rep("Diet", 6), rep("TissueMix", 19)))
  
  # Return final object
  list(
    Observation = Observation,
    Mock = Mock,
    N_pcr_mock = unique(input_mock_comm_RDS$N_pcr_mock),  # Get from input data
    NSpecies = nrow(sp_list),
    station_list = station_list,
    sp_list = sp_list
  )
}
```

Attempt at formatting fucntion attempt 2
```{r}
format_metabarcoding_data <- function(input_metabarcoding_RDS, input_mock_comm_RDS){
  require(tidyverse)
  
  Observation <- input_metabarcoding_RDS 
  
  # First create proper station indices for mock communities
  mock_community_index <- input_mock_comm_RDS %>%
    distinct(Community) %>%
    mutate(station_idx = row_number())
  
  Mock <- input_mock_comm_RDS %>% 
    # Join with station index mapping
    left_join(mock_community_index, by = "Community") %>%
    mutate(
      time = 1,    # for formatting compatibility
      creek = 1,
      biol = 1,
      # Use numeric station indices
      station = station_idx,
      # Convert species to sp_1, sp_2 format
      species = paste0("sp_", match(species, unique(species)))
    ) %>% 
    rename(Nreads = nReads) %>% 
    replace(is.na(.), 0) %>% 
    filter(start_prop > 0) %>% 
    filter(Nreads > 0) %>% # omit things that are absent from the mocks
    filter(!is.na(Nreads)) %>% 
    rename(b_proportion = start_prop) %>% 
    # Create Sample ID while keeping original columns
    unite(c(time, creek, Community, station, biol, tech), col = "Sample", sep = ".", remove = FALSE)
  
  # only keep species present in both mocks and observations
  keepSpecies <- intersect(Mock$species, Observation$species)
  Observation <- Observation %>% 
    filter(species %in% keepSpecies) 
  Mock <- Mock %>% 
    filter(species %in% keepSpecies)

  # index species to a common standard 
  sp_list <- data.frame(
    species = unique(c(Mock$species, Observation$species)),
    species_idx = match(unique(c(Mock$species, Observation$species)), 
                      unique(c(Mock$species, Observation$species)))
  )
  
  # Reindex and renormalize mock data
  Mock <- Mock %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      speciesname = species,
      species = species_idx
    ) %>% 
    group_by(station, tech, time, creek, biol) %>% 
    mutate(b_proportion = b_proportion/sum(b_proportion)) %>% 
    ungroup()
  
  # Process observation data
  Observation <- Observation %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      stationname = station,
      speciesname = species,
      species = species_idx
    ) %>% 
    mutate(station = case_when(
      stationname == "D1" ~ 1, stationname == "D2" ~ 2, stationname == "D3" ~ 3,
      stationname == "D4" ~ 4, stationname == "D5" ~ 5, stationname == "D6" ~ 6,
      stationname == "B_Skate" ~ 7, stationname == "MgArmSq" ~ 8, stationname == "BayShr" ~ 9,
      stationname == "MktSq" ~ 10, stationname == "GPO" ~ 11, stationname == "Ratfish" ~ 12,
      stationname == "RedOcto" ~ 13, stationname == "CoonStShrp" ~ 14, stationname == "SptPrawn" ~ 15,
      stationname == "PnkShrmp" ~ 16, stationname == "SpDog" ~ 17, stationname == "Lnose_skate" ~ 18,
      stationname == "RivLamp" ~ 19, stationname == "SqVAQ" ~ 20, stationname == "GhShrmp" ~ 21,
      stationname == "StubSq" ~ 22, stationname == "MusPolyWorm" ~ 23, 
      stationname == "BlkTigShrVAQ" ~ 24, stationname == "VAQShrimp" ~ 25,
      TRUE ~ NA_integer_
    )) %>% 
    group_by(station, time, creek, biol) %>% 
    mutate(
      tech = match(tech, unique(tech)),
      tot_reads = sum(Nreads)
    ) %>% 
    filter(tot_reads > 0) %>% 
    ungroup() %>% 
    dplyr::select(-tot_reads)
  
  # Create comprehensive station list
  station_list <- data.frame(
    station = 1:25,
    stationname = c(paste0("D", 1:6),
                   "B_Skate", "MgArmSq", "BayShr", "MktSq", "GPO", 
                   "Ratfish", "RedOcto", "CoonStShrp", "SptPrawn",
                   "PnkShrmp", "SpDog", "Lnose_skate", "RivLamp",
                   "SqVAQ", "GhShrmp", "StubSq", "MusPolyWorm",
                   "BlkTigShrVAQ", "VAQShrimp"),
    station_type = c(rep("Diet", 6), rep("TissueMix", 19)))
  
  return(
    list(
      Observation = Observation,
      Mock = Mock,
      N_pcr_mock = unique(input_mock_comm_RDS$N_pcr_mock), 
      NSpecies = nrow(sp_list),
      station_list = station_list,
      sp_list = sp_list
    )
  )
}
```

Modified formatting function 3
```{r}
# Modified formatting function that handles species matching better
format_metabarcoding_data <- function(input_metabarcoding_RDS, input_mock_comm_RDS) {
  require(tidyverse)
  
  Observation <- input_metabarcoding_RDS
  
  # Process Mock data FIRST - join station_idx before using it
  Mock <- input_mock_comm_RDS %>% 
    # First create station indices
    group_by(Community) %>% 
    mutate(station_idx = cur_group_id()) %>% 
    ungroup() %>%
    # Then proceed with other transformations
    rename(Nreads = nReads) %>%
    mutate(
      time = 1,
      creek = 1,
      biol = 1,
      station = station_idx,  # Now station_idx exists
      stationname = Community
    ) %>% 
    replace(is.na(.), 0) %>% 
    filter(start_prop > 0 & Nreads > 0 & !is.na(Nreads)) %>% 
    unite(c(time, creek, station, biol, tech), col = "Sample", sep = ".", remove = FALSE)
  
  # Keep species present in both datasets
  keepSpecies <- intersect(Mock$species, Observation$species)
  
  Observation <- Observation %>% 
    filter(species %in% keepSpecies)
  
  Mock <- Mock %>% 
    filter(species %in% keepSpecies)
  
  # Create species index (preserving original case)
  sp_list <- data.frame(
    species = unique(c(Mock$species, Observation$species)),
    species_idx = match(unique(c(Mock$species, Observation$species)), 
                      unique(c(Mock$species, Observation$species)))
  )
  
  # [Rest of your function remains exactly the same...]
  # Process Mock data with species index
  Mock <- Mock %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      speciesname = species,
      species = species_idx
    ) %>% 
    group_by(station, tech, time, creek, biol) %>% 
    mutate(b_proportion = start_prop/sum(start_prop)) %>% 
    ungroup()
  
  # Process Observation data
  Observation <- Observation %>% 
    left_join(sp_list, by = "species") %>% 
    mutate(
      stationname = station,
      speciesname = species,
      species = species_idx
    ) %>% 
    mutate(station = case_when(
      stationname == "D1" ~ 1, stationname == "D2" ~ 2, stationname == "D3" ~ 3,
      stationname == "D4" ~ 4, stationname == "D5" ~ 5, stationname == "D6" ~ 6,
      stationname == "B_Skate" ~ 7, stationname == "MgArmSq" ~ 8, stationname == "BayShr" ~ 9,
      stationname == "MktSq" ~ 10, stationname == "GPO" ~ 11, stationname == "Ratfish" ~ 12,
      stationname == "RedOcto" ~ 13, stationname == "CoonStShrp" ~ 14, stationname == "SptPrawn" ~ 15,
      stationname == "PnkShrmp" ~ 16, stationname == "SpDog" ~ 17, stationname == "Lnose_skate" ~ 18,
      stationname == "RivLamp" ~ 19, stationname == "SqVAQ" ~ 20, stationname == "GhShrmp" ~ 21,
      stationname == "StubSq" ~ 22, stationname == "MusPolyWorm" ~ 23, 
      stationname == "BlkTigShrVAQ" ~ 24, stationname == "VAQShrimp" ~ 25,
      TRUE ~ NA_integer_
    )) %>% 
    group_by(station, time, creek, biol) %>% 
    mutate(
      tech = match(tech, unique(tech)),
      tot_reads = sum(Nreads)
    ) %>% 
    filter(tot_reads > 0) %>% 
    ungroup() %>% 
    dplyr::select(-tot_reads)
  
  # Create station metadata
  station_list <- data.frame(
    station = 1:25,
    stationname = c(paste0("D", 1:6),
                   "B_Skate", "MgArmSq", "BayShr", "MktSq", "GPO", 
                   "Ratfish", "RedOcto", "CoonStShrp", "SptPrawn",
                   "PnkShrmp", "SpDog", "Lnose_skate", "RivLamp",
                   "SqVAQ", "GhShrmp", "StubSq", "MusPolyWorm",
                   "BlkTigShrVAQ", "VAQShrimp"),
    station_type = c(rep("Diet", 6), rep("TissueMix", 19)))
  
  # Return final object
  list(
    Observation = Observation,
    Mock = Mock,
    N_pcr_mock = unique(input_mock_comm_RDS$N_pcr_mock),
    NSpecies = nrow(sp_list),
    station_list = station_list,
    sp_list = sp_list
  )
}

```


```{r}
qmdata <- format_metabarcoding_data(input_metabarcoding_RDS, input_mock_comm_RDS)

obs <- qmdata

prop_check <- qmdata$Mock %>%
  group_by(station, tech) %>%
  summarise(
    sum_prop = sum(b_proportion),
    n_species = n(),
    .groups = 'drop'
  )

# Identify any problematic groups
prop_check %>% filter(abs(sum_prop - 1) > 0.001)

```
If no problems, continue to a sanity check

```{r}
p3 <- plot_proportion_comparison(qmdata$Mock,
                                 taxonomy_col = "speciesname",
                                 count_col = "Nreads",
                                 existing_prop_col = "propReads",
                                 group_cols = c("Community", "tech"),
                                 top_n_taxa = 12,
                                 plot_title = "Count vs Clac Proportions")
print(p3)
```
okay working as expected after transform

ORIGINAL RUNNING OF THE SCRIPT
```{r}
library(rstan)
library(MCMCpack) #for rdirichelet function
library(gridExtra)
library(unikn)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


N_pcr_cycles <- 39
MOCK <- qmdata$Mock   

##### ORIGINAL####
alrTransform <- function(MOCK){
  require(tidyverse)
  require(compositions)
  
  p_mock <- MOCK %>% 
    dplyr::select(species, station, tech, b_proportion) %>% 
    rename(tech_rep = tech) %>% 
    pivot_wider(names_from = species, values_from = b_proportion, values_fill = 1e-9) %>% 
    ungroup() 
  colnames(p_mock)[3:(length(unique(MOCK$species))+2)] <- paste0("alr_", 1:length(unique(MOCK$species)))
  
  p_mock <- alr(p_mock[,3:ncol(p_mock)]) %>% as.matrix() %>% as.data.frame()
  p_mock[,length(unique(MOCK$species))] <- 0  #add reference zero expressly
  names(p_mock)[length(unique(MOCK$species))] <- paste0("alr_", length(unique(MOCK$species)))
  
  p_mock <-  cbind(MOCK %>% dplyr::select(tech, station) %>% distinct(),
                   p_mock) %>% ungroup()
  names(p_mock)[1] <- "tech_rep"
  
  
  return(p_mock)
}

makeDesign <- function(obs, #obs is a named list with elements Observation, Mock, N_pcr_mock, sp_list
                       N_pcr_cycles){ #N_pcr_cycles is the number of PCR cycles in your experimental/enviro samples; currently a single value, could be made into a vector if this number varies
  #library(tidyverse)
  library(MCMCpack)
  library(compositions)
  library(rstan)
  library(dplyr)
  
  
  mock <- obs$Mock
  observed <- obs$Observation
  
  p_mock_all <- alrTransform(mock)
  
  mock_3 <- mock %>% 
    dplyr::select(species, station, tech, Nreads) %>% 
    ungroup() %>% 
    mutate(species = paste0("sp_", species)) %>% 
    pivot_wider(names_from = species, values_from = Nreads, values_fill = 0)
  N_pcr_mock <- rep(obs$N_pcr_mock, nrow(p_mock_all)) #assumes all have the same Npcr
  
  
  p_samp_all <- observed %>% 
    ungroup() %>% 
    unite(time, creek, station, biol, col = "station") %>% 
    dplyr::select(station, tech, species, Nreads) %>% 
    rename(tech_rep = tech) %>% 
    mutate(species = paste0("sp_", species)) %>% 
    arrange(species) %>% 
    group_by(station, tech_rep,species) %>% 
    dplyr::summarise(Nreads= sum(Nreads)) %>% 
    ungroup() %>% 
    pivot_wider(names_from = species, values_from = Nreads, values_fill = 0)
  N_pcr_samp <- rep(N_pcr_cycles, nrow(p_samp_all))
  
  ########################################################################
  #### Create data frames that can be read into Stan model
  ########################################################################
  
  NOM <- as.name(colnames(p_mock_all)[1])
  formula_a <- eval(NOM) ~ N_pcr_mock -1
  model_frame <- model.frame(formula_a, p_mock_all)
  model_vector_a_mock <- model.matrix(formula_a, model_frame) %>% as.numeric()
  N_pcr_mock_small <- cbind(N_pcr_mock, p_mock_all) %>%  filter(tech_rep == 1) %>% pull(N_pcr_mock)
  formula_b <- eval(NOM) ~ N_pcr_mock_small -1
  model_frame <- model.frame(formula_b, p_mock_all%>% filter(tech_rep==1))
  model_vector_a_mock_small <- model.matrix(formula_b, model_frame) %>% as.numeric()
  
  N_obs_mock       <- nrow(p_mock_all)
  
  # unknown communities second
  # species compositions (betas)
  
  NOM <- as.name(colnames(p_samp_all)[1])    
  
  p_samp_all$station <- as.factor(p_samp_all$station)
  N_station = length(unique(p_samp_all$station))
  p_samp_all$tech_rep <- as.factor(p_samp_all$tech_rep)
  if(N_station == 1){
    formula_b <- eval(NOM) ~ 1  
  } else {
    formula_b <- eval(NOM) ~ station
  }
  
  model_frame <- model.frame(formula_b, p_samp_all)
  model_matrix_b_samp <- model.matrix(formula_b, model_frame)
  
  # choose a single representative for each station to make predictions to
  model_frame <- model.frame(formula_b, p_samp_all %>% filter(tech_rep==1))
  model_matrix_b_samp_small <- model.matrix(formula_b, model_frame)
  
  # efficiencies (alpha)
  formula_a <- eval(NOM) ~ N_pcr_samp -1
  model_frame <- model.frame(formula_a, p_samp_all)
  model_vector_a_samp <- model.matrix(formula_a, model_frame) %>% as.numeric()
  N_pcr_samp_small <- cbind(N_pcr_samp, p_samp_all) %>% filter(tech_rep == 1) %>% pull(N_pcr_samp)
  formula_b <- eval(NOM) ~ N_pcr_samp_small -1
  
  model_frame <- model.frame(formula_b, p_samp_all %>% filter(tech_rep==1))
  model_vector_a_samp_small <- model.matrix(formula_b, model_frame) %>% as.numeric()
  
  #counters 
  N_obs_samp_small <- nrow(model_matrix_b_samp_small)
  N_obs_samp <- nrow(p_samp_all)
  N_b_samp_col <- ncol(model_matrix_b_samp)
  
  
  #### Make Stan objects
  
  stan_data <- list(
    N_species = ncol(p_samp_all)-2,   # Number of species in data
    N_obs_samp = nrow(p_samp_all), # Number of observed community samples and tech replicates ; this will be Ncreek * Nt * Nbiol * Ntech * 2 [for upstream/downstream observations]
    N_obs_mock = nrow(p_mock_all), # Number of observed mock samples, including tech replicates
    N_obs_samp_small = nrow(p_samp_all %>% filter(tech_rep == 1)), # Number of unique observed community samples ; this will be Ncreek * Nt * Nbiol * 2 [for upstream/downstream observations]
    
    # Observed data of community matrices
    sample_data = p_samp_all %>% dplyr::select(contains("sp")),
    sample_vector = unique(p_samp_all$station),
    mock_data   = mock_3 %>% dplyr::select(contains("sp")),
    sp_list = obs$sp_list,
    
    # True proportions for mock community
    #mock_true_prop = p_mock_all %>% dplyr::select(contains("sp")),
    alr_mock_true_prop = p_mock_all %>% dplyr::select(contains("alr")),
    
    # vectors of PCR numbers
    N_pcr_samp = N_pcr_samp,
    N_pcr_mock = N_pcr_mock,
    
    # Design matrices: field samples
    N_b_samp_col = N_b_samp_col,
    model_matrix_b_samp = model_matrix_b_samp,
    model_matrix_b_samp_small = as.array(model_matrix_b_samp_small),
    model_vector_a_samp = model_vector_a_samp,
    model_vector_a_samp_small = as.array(model_vector_a_samp_small),
    
    # Design matrices: mock community samples
    model_vector_a_mock = as.array(model_vector_a_mock),
    
    # Priors
    alpha_prior = c(0,0.5),  # normal prior
    beta_prior = c(0,5),    # normal prior
    tau_prior = c(1,2)   # gamma prior
  )
  
  return(stan_data)
  
}


stan_metabarcoding_data <- makeDesign(qmdata, N_pcr_cycles = 39)
stan_metabarcoding_data$alr_mock_true_prop
stan_metabarcoding_data$mock_data
stan_metabarcoding_data$sample_vector

```
Modified model running script to handle changes that are leading to crazy issues like not setting the reference species to an appropriate one
```{r}
library(rstan)
library(MCMCpack) #for rdirichelet function
library(gridExtra)
library(unikn)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


N_pcr_cycles <- 39
MOCK <- qmdata$Mock 
ref_species = 5

```

Modified ALR transform
```{r}
##### Modified first ####
# Modify the alrTransform function to ensure consistent row naming - set reference as 5 = Actinopteri
alrTransform <- function(MOCK, ref_species) {
  require(tidyverse)
  require(compositions)
  
  # Ensure complete data with explicit factor levels
  mock_complete <- MOCK %>%
    complete(Sample = unique(Sample), 
             tech = unique(tech),
             species = unique(species),
             fill = list(Nreads = 0, b_proportion = 1e-9))
  
  # Create proportions matrix without row names
  p_mock <- mock_complete %>% 
    dplyr::select(Sample, tech, species, b_proportion) %>%
    pivot_wider(names_from = species, values_from = b_proportion, values_fill = 1e-9) %>%
    arrange(Sample, tech)  # Ensure consistent ordering
    
  # Apply ALR transformation
  alr_matrix <- alr(as.matrix(p_mock[,-(1:2)]), 
                   ivar = which(names(p_mock) == as.character(ref_species)))
  
  # Return as simple data frame without row names
  as.data.frame(alr_matrix) %>% 
    set_names(paste0("alr_", setdiff(1:12, ref_species))) %>% 
    mutate(Sample = p_mock$Sample, tech = p_mock$tech) %>%
    dplyr::select(Sample, tech, everything())
}

```
We are going to split up the functions because as written in one giant script is making troubleshooting hard, later we can make them all into one Functions.R but for now let's keep them separate

```{r}
makeDesign <- function(obs, N_pcr_cycles) {
  library(MCMCpack)
  library(compositions)
  library(rstan)
  library(dplyr)
  
  mock <- obs$Mock
  observed <- obs$Observation
  
  # Debug: Print mock data structure
  cat("\n=== Mock Data Structure ===\n")
  print(distinct(mock %>% select(Sample, tech)))
  
  n_communities <- length(unique(mock$Sample))
  n_tech_reps <- length(unique(mock$tech))
  expected_mock_rows <- n_communities * n_tech_reps
  
  cat("\n=== Expected Dimensions ===\n")
  cat("Communities:", n_communities, "| Tech reps:", n_tech_reps, "| Expected rows:", expected_mock_rows, "\n")
  
  # Calculate expected number of mock samples
  n_communities <- length(unique(mock$Community))
  n_tech_reps <- length(unique(mock$tech))
  expected_mock_rows <- n_communities * n_tech_reps
  
   # Process mock data with completion check
  mock_3 <- mock %>%
    complete(Sample = unique(Sample),
             tech = unique(tech),
             species = unique(species),
             fill = list(Nreads = 0)) %>%
    mutate(species = paste0("sp_", species)) %>%
    pivot_wider(names_from = species, values_from = Nreads, values_fill = 0)
  
  cat("\n=== Actual Dimensions ===\n")
  cat("mock_3 rows:", nrow(mock_3), "\n")
  
  # ALR transformation
  p_mock_all <- alrTransform(mock) %>%
    arrange(Sample, tech) %>%
    mutate(tech_rep = tech)
  
  # Enhanced validation
  if(nrow(mock_3) != expected_mock_rows) {
    cat("\n=== Missing Combinations ===\n")
    all_combos <- expand.grid(
      Sample = unique(mock$Sample),
      tech = unique(mock$tech)
    )
    existing_combos <- distinct(mock, Sample, tech)
    missing <- anti_join(all_combos, existing_combos, by = c("Sample", "tech"))
    print(missing)
    
    stop("Mock data incomplete. Missing ", nrow(missing), " sample/tech combinations")
  }
    
  # Flexible PCR cycles
  N_pcr_mock <- rep(obs$N_pcr_mock, expected_mock_rows)

  
  # Process environmental samples (original code unchanged)
  p_samp_all <- observed %>% 
    ungroup() %>% 
    unite(time, creek, station, biol, col = "station") %>% 
    dplyr::select(station, tech, species, Nreads) %>% 
    rename(tech_rep = tech) %>% 
    mutate(species = paste0("sp_", species)) %>% 
    arrange(species) %>% 
    group_by(station, tech_rep, species) %>% 
    dplyr::summarise(Nreads = sum(Nreads), .groups = 'drop') %>% 
    pivot_wider(names_from = species, values_from = Nreads, values_fill = 0)
  
  N_pcr_samp <- rep(N_pcr_cycles, nrow(p_samp_all))
  
  ########################################################################
  #### Create data frames that can be read into Stan model
  ########################################################################
  
  NOM <- as.name(colnames(p_mock_all)[1])
  formula_a <- eval(NOM) ~ N_pcr_mock -1
  model_frame <- model.frame(formula_a, p_mock_all)
  model_vector_a_mock <- model.matrix(formula_a, model_frame) %>% as.numeric()
  N_pcr_mock_small <- cbind(N_pcr_mock, p_mock_all) %>%  filter(tech_rep == 1) %>% pull(N_pcr_mock)
  formula_b <- eval(NOM) ~ N_pcr_mock_small -1
  model_frame <- model.frame(formula_b, p_mock_all%>% filter(tech_rep==1))
  model_vector_a_mock_small <- model.matrix(formula_b, model_frame) %>% as.numeric()
  
  N_obs_mock       <- nrow(p_mock_all)
  
  # unknown communities second
  # species compositions (betas)
  
  NOM <- as.name(colnames(p_samp_all)[1])    
  
  p_samp_all$station <- as.factor(p_samp_all$station)
  N_station = length(unique(p_samp_all$station))
  p_samp_all$tech_rep <- as.factor(p_samp_all$tech_rep)
  if(N_station == 1){
    formula_b <- eval(NOM) ~ 1  
  } else {
    formula_b <- eval(NOM) ~ station
  }
  
  model_frame <- model.frame(formula_b, p_samp_all)
  model_matrix_b_samp <- model.matrix(formula_b, model_frame)
  
  # choose a single representative for each station to make predictions to
  model_frame <- model.frame(formula_b, p_samp_all %>% filter(tech_rep==1))
  model_matrix_b_samp_small <- model.matrix(formula_b, model_frame)
  
  # efficiencies (alpha)
  formula_a <- eval(NOM) ~ N_pcr_samp -1
  model_frame <- model.frame(formula_a, p_samp_all)
  model_vector_a_samp <- model.matrix(formula_a, model_frame) %>% as.numeric()
  N_pcr_samp_small <- cbind(N_pcr_samp, p_samp_all) %>% filter(tech_rep == 1) %>% pull(N_pcr_samp)
  formula_b <- eval(NOM) ~ N_pcr_samp_small -1
  
  model_frame <- model.frame(formula_b, p_samp_all %>% filter(tech_rep==1))
  model_vector_a_samp_small <- model.matrix(formula_b, model_frame) %>% as.numeric()
  
  #counters 
  N_obs_samp_small <- nrow(model_matrix_b_samp_small)
  N_obs_samp <- nrow(p_samp_all)
  N_b_samp_col <- ncol(model_matrix_b_samp)
  
  
  #### Make Stan objects
  
  # Create Stan data list with flexible dimensions
  stan_data <- list(
    N_species = length(unique(mock$species)),
    N_obs_samp = nrow(p_samp_all),
    N_obs_mock = expected_mock_rows,  # Now flexible
    N_obs_samp_small = nrow(p_samp_all %>% filter(tech_rep == 1)),
    
    # Data matrices
    sample_data = as.matrix(p_samp_all %>% select(starts_with("sp_"))),
    mock_data = as.matrix(mock_3 %>% select(starts_with("sp_"))),
    alr_mock_true_prop = as.matrix(p_mock_all %>% select(starts_with("alr_"))),
    
    N_pcr_samp = N_pcr_samp,
    N_pcr_mock = N_pcr_mock,
    
    N_b_samp_col = N_b_samp_col,
    model_matrix_b_samp = model_matrix_b_samp,
    model_matrix_b_samp_small = as.array(model_matrix_b_samp_small),
    model_vector_a_samp = model_vector_a_samp,
    model_vector_a_samp_small = as.array(model_vector_a_samp_small),
    model_vector_a_mock = as.array(model_vector_a_mock),
    
    alpha_prior = c(0,0.5),
    beta_prior = c(0,5),
    tau_prior = c(1,2)
  )
  
  # Add metadata for debugging
  attr(stan_data, "mock_info") <- list(
    n_communities = n_communities,
    n_tech_reps = n_tech_reps,
    expected_rows = expected_mock_rows
  )
  return(stan_data)
  
}
```

Running the nested functions
```{r}
stan_metabarcoding_data <- makeDesign(qmdata, N_pcr_cycles = 39)
stan_metabarcoding_data$alr_mock_true_prop
stan_metabarcoding_data$mock_data
stan_metabarcoding_data$sample_vector

```




okay now we can move forward with the next steps in the stan model and QM model

From Zack:
```{r}

QM_likelihood <- function(stanmodelname, stan_metabarcoding_data){
  M <- stan_model(stanmodelname)
  
  stanOpt <- optimizing(M, data=stan_metabarcoding_data, iter=30000,draws=0,
                        verbose=T,  
                        tol_param=1e-40,
                        algorithm="LBFGS",
                        hessian = TRUE)
  
  MLest <- stanOpt$par[grep("int_samp_small", names(stanOpt$par))] %>%
    matrix(ncol = stan_metabarcoding_data$N_species) %>% 
    as.data.frame()
  names(MLest) <- stan_metabarcoding_data$sp_list$species
  rownames(MLest) <- stan_metabarcoding_data$sample_vector
  ML_a <- stanOpt$par[grep("alpha\\[", names(stanOpt$par))]  
  ML_a <- data.frame("alpha_est" = ML_a, 
                     "species" = stan_metabarcoding_data$sp_list$species)
  
  
  return(list(
    ML_modelfit = stanOpt,
    ML_estimates = MLest,
    ML_alpha_est = ML_a
  ))
  
}

ML_out <- QM_likelihood(here("scripts","models","quant_metabar_rosetta_noSampleEta.stan"), stan_metabarcoding_data)

write_rds(ML_out, here("intermediate-files","Test2_CombFishMock_20250818_ML_out.RDS"))

```

So i think it kinda failed, and the outputs are very odd, so let's look at them visually

```{r}
library(tidyverse)

# Define station mapping (as you provided)
station_mapping <- data.frame(
  station = 1:25,
  station_name = c("D1", "D2", "D3", "D4", "D5", "D6",
                   "B_Skate", "MgArmSq", "BayShr", "MktSq", "GPO", 
                   "Ratfish", "RedOcto", "CoonStShrp", "SptPrawn",
                   "PnkShrmp", "SpDog", "Lnose_skate", "RivLamp",
                   "SqVAQ", "GhShrmp", "StubSq", "MusPolyWorm",
                   "BlkTigShrVAQ", "VAQShrimp")
)

library(tidyverse)

# Check sampletype values first
sample_types <- ML_out$ML_estimates %>%
  rownames_to_column("sample") %>%
  separate(col = sample, into = c("time", "sampletype", "station_num", "biol"), sep = "_") %>%
  pull(sampletype) %>%
  unique()
print(sample_types)  # Verify if "PvcF" or "PvCF" exists

# Proceed with corrected filter
plot_data_CF <- ML_out$ML_estimates %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "species", values_to = "proportion") %>%
  separate(col = sample, into = c("time", "sampletype", "station_num", "biol"), 
           remove = FALSE, sep = "_") %>% 
  mutate(station_num = as.numeric(station_num)) %>%
  left_join(station_mapping, by = c("station_num" = "station")) %>%
  filter(proportion > 0.001) %>%
  filter(sampletype == "PvCF")  # Adjusted to match actual case

# Check if plot_data has rows
if (nrow(plot_data_CF) == 0) {
  stop("No data left after filtering! Check sampletype/stations.")
}


plot_data_TM <- ML_out$ML_estimates %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "species", values_to = "proportion") %>%
  separate(col = sample, into = c("time", "sampletype", "station_num", "biol"), 
           remove = FALSE, sep = "_") %>% 
  mutate(station_num = as.numeric(station_num)) %>%
  left_join(station_mapping, by = c("station_num" = "station")) %>%
  filter(proportion > 0.001) %>%
  filter(sampletype == "TM")  # Adjusted to match actual case

# Check if plot_data has rows
if (nrow(plot_data_TM) == 0) {
  stop("No data left after filtering! Check sampletype/stations.")
}


# Plot
ggplot(plot_data_CF, aes(x = biol, fill = species, y = proportion)) +
  geom_col(position = "stack", width = 0.8) +
  labs(x = "Biological Replicate", y = "Proportion in sample", fill = "Species") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    legend.position = "bottom"
  ) +
  facet_grid(sampletype ~ station_name + time, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 3))



ggsave(here("Figures","20250815_ctrlfeeding_proportions_after_qm_ML.png"), width = 18, height = 10, units = "cm")


```

Okay, well there is a lot of noise, and it is hard to visualise, but there is def stuff wrong, we need to clean up the data potentially more, or split the TissueMixtures from the Controlled feeding study to run separately... Something is wrong, and the model didn't really finish running well from the looks of it

```{r}
# Plot
ggplot(plot_data_TM, aes(x = biol, fill = species, y = proportion)) +
  geom_col(position = "stack", width = 0.8) +
  labs(x = "Biological Replicate", y = "Proportion in sample", fill = "Species") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    legend.position = "bottom"
  ) +
  facet_grid(sampletype ~ station_name + time, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 3))

```
After including all the species and having both the CF and TM samples in there, it looks like the model is really not working to handle the zeros as it should, mammal and spiny dogfish are wildly over represented... 


okay let's run some checks:
```{r}
# Check species lists
cat("Mock species:", unique(qmdata$Mock$species), "\n")
cat("Observation species:", unique(qmdata$Observation$species), "\n")

# Check dimensions
cat("Mock data dimensions:", dim(stan_metabarcoding_data$mock_data), "\n")
cat("Sample data dimensions:", dim(stan_metabarcoding_data$sample_data), "\n")

# Check ALR proportions
print("First few rows of ALR transformed mock proportions:")
print(head(stan_metabarcoding_data$alr_mock_true_prop))

# Check design matrices
print("Design matrix for samples:")
print(head(stan_metabarcoding_data$model_matrix_b_samp))
print("PCR cycles vector:")
print(head(stan_metabarcoding_data$model_vector_a_samp))
```

First major issue flagged is that Actinopteri is not being set as the reference, Oegopsida is which isn't in all the samples like Actinopteri is. This is showing up as an obvious problem with the ALR transform step.

```{r}
####################################################################
# Run Bayesian QM model
####################################################################

bayes_out <- QM_bayes("/Users/zgold/Documents/GitHub/quantitative_salmon_culverts/Scripts/functions/quant_metabar_rosetta_noSampleEta.stan", stan_metabarcoding_data)

saveRDS(bayes_out, here("analysis","20231210_bayesout_salmon_only.RDS"))

summaryout <- summary(bayes_out$Bayes_modelfit)$summary
write.csv(summaryout, here("analysis","20231211_bayesout_salmon_eul_only_summary.csv"))

bayes_out <- readRDS( here("analysis","20231210_bayesout_salmon_only.RDS"))

bayes_out$Bayes_alpha_est
bayes_out$Bayes_estimates %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "species") -> tibble_bayes

bayes_out$Bayes_25ci %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "species") -> tibble_bayes_25

bayes_out$Bayes_75ci %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "species") -> tibble_bayes_75

tibble_bayes$Bayes_25ci <- tibble_bayes_25$value
tibble_bayes$Bayes_75ci <- tibble_bayes_75$value

bayes_out$Bayes_modelfit
pars2 <- rstan::extract(bayes_out$Bayes_modelfit, par = "int_samp_small")

species_mapping <- stan_metabarcoding_data$sp_list
Stations <-  stan_metabarcoding_data$sample_vector %>% as.tibble()
Stations %>% 
  mutate(., station_idx =row_number())-> Stations

pars2$int_samp_small %>% 
    as_tibble() %>% 
    pivot_longer(., cols = `1.1`:`197.6`, names_to="station.species", values_to = "est") %>%
    separate(station.species, into = c("station_idx","species_idx"), remove = F, sep="\\.") %>%
    mutate(., station_idx = as.numeric(station_idx),
           species_idx = as.numeric(species_idx)) %>%
     left_join(species_mapping) %>%
       left_join(Stations) -> int_samp_small_tibble

saveRDS(int_samp_small_tibble, here("analysis","int_samp_small_tibble.RDS"))

```

